---
title: "Regression Model - Course Proj."
author: "thecapacity"
date: "February 22, 2015"
output: html_document
---

### Summary

Using a linear regression model, based on the `mtcars` dataset provided by R, our model found that a car with manual transmission is more effieicnt per gallon than a car with automatic transmission ( ```r round(mean( mtcars[ mtcars$am == 1,]$mpg ),2)``` vs. ```r round(mean( mtcars[ mtcars$am == 0,]$mpg ),2)```). This estimate, that `am` represents two distinct populations, was validated with a `p.value` of ```r round(t.test(mpg ~ am, data=mtcars)$p.value,4)```. It was also discovered that the weight (`wt`) also plays an importat role as well. We only superficially explored interactions across variables. Based on an assumption 'power' indicators may be correlated with `wt` weight, such as `cyl` or engine shape `vs`, the potential exists for further exploration, including whether `hp`  or `disp`is a better predictor than `cyl`.

### Analysis

I am interested in exploring the relationship between a set of variables and and an efficiency outcome, in miles per gallon (MPG), in the cars data collected. **We are particularly interested in the following two questions:**

1. *Is an automatic or manual transmission better for MPG?* - **Answer: `Manual`**

2. *Quantify the MPG difference between automatic and manual transmissions.* - **Answer: ```r round( mean( mtcars[ mtcars$am == 1,]$mpg ) - mean( mtcars[ mtcars$am == 0,]$mpg ),2)```**

#### Approach

*Note: Pre-processing is minimal, and is included in the markdown file - **but not shown** - to reduce overall page count.*

```{r, echo=FALSE}
# Ensure pristine working environment
rm(list=ls())

library(knitr)
opts_chunk$set(echo = TRUE, fig.path="figure/", dev="png")
options(scipen = 1) # Turn off scientific notations for numbers

data(mtcars)
mtcars$am = as.factor(mtcars$am)
mtcars$cyl = as.factor(mtcars$cyl)
mtcars$vs = as.factor(mtcars$vs)
mtcars$gear = as.factor(mtcars$gear)
mtcars$carb = as.factor(mtcars$carb)
```

Based on my *'expertise'*, I believe cylinders `cyl` would be another key factor to consider in addition to `am` type. 

**Model 1:** However, we begin with the most parsimonious model then invetgiate other variables:
```{r, results='hide'}
fit1 <- lm(mpg ~ am, mtcars) #`am1` represnets a manual transmission
summary(fit1) #output hidden for pagecount
```

The model shows that `am` is significant at `0.001` level, with a `t value` of ```r round(summary(fit1)$coef[6],3)```, in influencing `mpg`. However, the `Adjusted R-squared` value is only ```r round(summary(fit1)$adj.r.squared,4)```, which means that the model can only predict about `33.85%` of the `mpg` variable. **Clearly some improvement is desired.** The residual deiviance (```r round(deviance(fit1),2)```) seems high, but **we will save more thorough review once we feel confidant in the model.**

**Model 2:** We then felt it would be useful to compare predicting `mpg` as a function of all variables:
```{r, results='hide'}
fit2 <- lm(mpg ~ . - 1, mtcars) #origin not included to make factor comparisons easier
summary(fit2) #output hidden for pagecount
```

Here we see that the estimated coefficient for `wt` of ```r round(summary(fit2)$coef[7],2)```, with a signficance code of `0.1`, `hp` also has similar significance, but only a ```r round(summary(fit2)$coef[5],2)``` coefficient. **Also interesting, is that impact by number of cylinders seems to be counter to an expected result**, i.e. `cyl4` and `cyl8` show similar coefficients, though all `cyl` values have large margins for `Std. Error`. This impact seems counter to our intuition, and **we believe this is an important indicater for future exploration**. _In addition the `R-squared` value of ```r round(summary(fit2)$r.squared,3)``` too high to be reasonable_.

**Model 3:** *Next, we used a model using `am`, `wt`, and `cyl`.* This was selected due to the factors mentioned above, excluding `hp` because of a ```r round(cor(mtcars$wt, mtcars$hp),3)``` corrlation with `wt`, and the small coefficient mentioned earlier.
```{r, results='hide'}
fit3 <- lm(mpg ~ am + wt + cyl, mtcars) 
summary(fit3) #output hidden for pagecount
```

From this model, we saw that `wt` and `cyl` were significant at `0.01` levels. **We also saw a more intuitive relationship between `mpg` and `cyl`**, specifically that mpg decreases more for `cyl8` than `cyl4`. Interestingly, `am` was not deemed significant; we assume this may be the result of an interaction between `am` likely with `wt`.

Therefore our final model contains these factors and the interaction (`wt`:`am`):
```{r}
fit4_final <- lm(mpg ~ am + cyl + wt + am:wt, mtcars)
summary(fit4_final) #$coef #abbreviated output for pagecount
```

All factors are significant and align with relationships automotive enthuiasts would expect; that MPG goes up for manual cars (`am == 1`) and down as `cyl` increases. **With a residual variance of ```r summary(fit4_final)$sigma``` and an `Adjusted R-squared` of ```r summary(fit4_final)$adj.r.squared``` this model seems to be quite accurate and balances simplicity**. A confidence interval `confint(fit4_final)` suggests some challenges; because the `cyl6` interval includes `0`. **One followup would be whether `cyl` interacts with other variables like `wt`.**

The appendix plots support the expected residual behavior. There is no consistent *"Residuals vs Fitted"* pattern, and the *"Normal Q-Q"* plot shows points close along the line, suggesting normal distribution. The *"Scale-Location"* plot suggests a reasonable constant variance assumption and *"Residuals vs Leverage"* shows no obvious outliers. **Note:** `echo = FALSE` was added to code below to meet the requirement that the appendix only include figures.

### Appendix

#### Best Fit Plot
```{r, echo=FALSE}
par(mfrow= c(2,2))
plot(fit4_final)
```

#### Residuals vs. X's in Final Model
```{r, echo=FALSE}
par(mfrow= c(1,3))
plot(mtcars$am, resid(fit4_final), main="Residuals by Transmission")
plot(mtcars$cyl, resid(fit4_final), main="Residuals by # of Cylinders")
plot(mtcars$wt, resid(fit4_final), main="Residuals by Weight")
```

#### Plot of `mpg` vs our final key factors, `am` and `cyl`.
```{r, echo=FALSE}
par(mfrow= c(1,3))
plot(mtcars$am, mtcars$mpg, main="MPG by Transmission")
plot(mtcars$cyl, mtcars$mpg, main="MPG by # of Cylinders")
plot(mtcars$wt, mtcars$mpg, main="MPG by Weight")
```
