---
title: "ToothGrowth Exploratory Analysis"
author: "thecapacity"
date: "January 24, 2015"
output: html_document
---

## Overview

This is the project file for my Peer Assignment of the Statistical Inference class.

I will seek to analyze the ToothGrowth data in the R datasets package, by:

1. Loading the ToothGrowth data

2. Performing some basic exploratory data analyses

3. Providing a basic summary of the data.

4. Using confidence intervals and/or hypothesis tests to compare tooth growth by ```supp``` and ```dose```. *Note I will only use techniques from class*

5. Stating my conclusions and the assumptions needed for those conclusions.

I will create a report to answer the questions. Given the nature of the series, ideally knitr will be used to create the reports and convert to a pdf. The pdf report will be **no more than 6 pages total including supporting material** if needed (code, figures, etcetera).

This will also be published at: http://rpubs.com/thecapacity/StatInf_Proj2

## Approach

I will plan to ensure I cover the following activities:

* Perform an exploratory data analysis with at least a single plot or table highlighting basic features of the data?

* Perform some relevant confidence intervals and/or tests?

* Review the results of the tests and/or intervals and interpret them in the context of the problem?

* Describe the assumptions needed for conclusions?

### Load Data
Here we focus on loading the ToothGrowth data from the R datasets package, and provide some initial information on the data size and structure.

```{r}
# Load the Data

```

### Analysis

Using confidence intervals and/or hypothesis tests to compare tooth growth by ```supp``` and ```dose```. I will perform some basic exploratory data analyses that will be collected and visualized, to be summarized in the next section.

This will include:

* At least a single plot or table highlighting basic features of the data

* A relevant confidence intervals and/or other potential tests

**Note I will only use techniques from class.**

```{r}
# Begin Basic Exploratory Analysis
```

### Summary
The following section provides a basic summary of the data, including the tests and/or intervals used, along with contextual interpretation.

### Conclusions

The following discussion focused on restating my conclusions and the assumptions needed for those conclusions and further describing assumptions underlying the analysis and summary.
